The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2016-04-10 script/en/chapters/ellipticcurves.tex
CT Book ed 12: Corrected some typos and updated some parts in the chapters about bitciphers and ECC.

\newpage
\hypertarget{ellcurve}{}
\chapter{Elliptic Curves}
\label{Chapter_EllipticCurves}
\index{elliptic curve}
(Filipovics B. / B\"uger M. / Esslinger B. / Oyono R., April 2000,
Updates: Dec. 2001, June 2002, Mar. 2003, Nov. 2009, Aug. 2013)

\section{Elliptic curve cryptography -- a high-performance substitute for RSA?}
\index{performance}
\label{ECAlternative}

In many business sectors secure and efficient data transfer is essential.
In particular, the RSA algorithm is used in many applications. Although the
security of RSA is beyond doubt, the evolution in computing power has
caused a growth in the necessary key length.  Today, 1024-bit RSA keys are
standard, but the GISA\index{GISA} (German Information Security Agency)
recommends the usage of 2048-bit keys from 2006 on (compare
section~\ref{SecurityRSA}).  The fact that most chips on smart cards cannot
process keys extending 1024 bit shows that there is a need for
alternatives. Elliptic curve cryptography (ECC) can be such an
alternative in the area of asymmetric cryptography.

@@

In addition, a digital signature can be processed 10-times faster with ECC
than with RSA.  However, verification of a given signature is still more
efficient with RSA than with ECC. Refer to
figure~\ref{ThousandBitMultiplications} (source: Dr.~J.\ Merkle, Elliptic
Curve Cryptography Workshop, 2001) for a comparison.  The reason is that
RSA public keys can be chosen relatively small as long as the secret key is
long enough.

@@

Although we see no need to replace RSA in any application today\footnote{%
Current information about the security of the RSA algorithm can be found in 
chapter \ref{SecurityRSA}.}, one should
take the usage of ECC-based tools into consideration whenever a new system
is set up --- in particular, when the tool should be available beyond 2005\footnote{%
Compare the recommendation of GISA: ``Fitting Crypto Algorithms'' from October 24th, 2002.
}.

\section{Elliptic curves -- history}

@@

With a free translation, using the denotation of modern mathematics, this means: \\
No positive whole numbers $x, y$ and $z$ greater than zero exist such that $x^n +y^n = z^n$ for $n>2$. I have found an amazing proof of this fact, but there is
too little space within the confines of this book to include it.

@@

\section{Elliptic curves -- mathematical basics}

@@

\subsection{Groups}

@@

\subsection{Fields}

@@

Fields may contain an infinite number of elements (e.g.\ the field of real numbers). They are called {\em infinite} fields. In contrast we call a field
{\em finite}, if it contains only a finite number of elements (e.g.\ ${\mathbb Z}_p = \{0, 1, 2, 3, \cdots, p-1\}$
, where $p$ is a prime. ${\mathbb Z}_p$ with addition mod $p$ and multiplication
mod $p$).
\index{field!characteristic}
\paragraph*{Characteristic of a field}
Let $K$ be a field and $1$ be the neutral element of $K$ with
respect to the multiplicative operation ``$\cdot$''. Then the characteristic of $K$ is said to be the order of $1$ with respect to the additive operation. This means that the characteristic of $K$ is the smallest positive integer $n$ such that
$$ \underbrace{1+1+\dots+1}_{\hbox{$n$ times}} =0 .
$$
If there is no such $n$, i.e. if $1+1+\dots+1\ne 0$ no matter how many $1$s we add, then we call $K$ a field
with characteristic $0$.

@@

In cryptography elliptic curve are a useful tool. Such curves are described by some equation. A detailed analysis has shown that curves of the form\footnote{This curve is given by the zeros of a {\it polynomial}\index{polynomial} $F$ of degree three in three variables. In general, expressions of the form
$P=\sum_{i_1,\dots,i_n\in\N_0} a_{i_1\dots i_n} x_1^{i_1}\dots x_n^{i_n}$ with coefficients $a_{i_1\dots i_n}\in K$ are called polynomials in $n$ variables $x_1,\dots,x_n$ with underlying field $K$, if ${\rm deg\,} P:=\max\{i_1+\dots +i_n: a_{i_1\dots i_n}\ne 0\}$ is finite, i.e. the sum has only finitely many non-zero terms (monomials). The sum of the exponents of the variables of each term of the sum is at most $3$, at least one term of the sum has a single variable with $3$ as value of the according exponent.}
\begin{equation}
 F(x_1,x_2,x_3)=-x_1^3+x_2^2x_3+a_1x_1x_2x_3-a_2x_1^2x_3+a_3x_2x_3^2-a_4x_1x_3^2-a_6x_3^3=0,
\label{eccbasisgleichung}
\end{equation}
are especially useful. The variables $x_1,x_2,x_3$ and parameters $a_1,\dots,a_4,a_6$ are elements of a given field $K$, which has certain properties that are make it useful from the cryptographic point of view. The underlying field $K$ might be the well known field of real numbers or some finite field (see last section).
In order to obtain a cure that is useful for cryptography, the parameters have to be chosen in a way that the following conditions hold
$$ \frac{\partial F}{\partial x_1}\ne 0, \quad \frac{\partial F}{\partial x_2}\ne 0, \quad
\frac{\partial F}{\partial x_3}\ne 0 .
$$
We identify points on the curve that can be derived from each over by multiplying each component with some scalar. This makes sense since $(x_1,x_2,x_3)$ solves (\ref{eccbasisgleichung}) if and only if $\alpha (x_1,x_2,x_3)$ ($\alpha\ne 0$) does. Formally, this means that we consider classes of equivalent points instead of single points, where points are called equivalent if one is the scalar multiple of the other one.
\\ If we put $x_3=0$ in the basic equation (\ref{eccbasisgleichung}), then this equation collapses to $-x_1^3=0$, leading to $x_1=0$. Thus, the equivalence class which includes the element $(0,1,0)$ is the only one that contains a point with $x_3=0$. For all points on the curve that are not equivalent to $(0,1,0)$, we may apply the following transformation
$$ K\times K\times (K\setminus\{0\})\ni (x_1,x_2,x_3) \mapsto (x,y):=\left( \frac{x_1}{x_3}, \frac{x_2}{x_3}\right) \in K\times K \, ,
$$
which reduces the number of variables to two instead of three. We note that the
basic equation (\ref{eccbasisgleichung}) $F(x_1,x_2,x_3)=0$  was chosen in a
way that this transformation leads to the famous so-called
Weierstrass-Equation\footnote{Karl Weierstrass\index{Weierstrass, Karl}, 31.10.1815$-$19.12.1897, German
mathematician, famous for his rigorous formal approach to mathematics.} holds
\begin{equation}
 y^2+a_1xy+a_3y = x^3+a_2x^2+a_4x+a_6 \, .
\label{ell}
\end{equation}
Since all but one point (i.e. equivalence class) of the elliptic curve can be described using equation (\ref{ell}), this equation is often called the elliptic equation, and its solutions written as
$$ {\bf E} = \left\{(x,y)\in K\times K \, |\, y^2+a_1xy+a_3y = x^3+a_2x^2+a_4x+a_6  \right\} \cup \{{\cal O} \}.
$$
Here, ${\cal O}$ represents the point $(0,1,0)$ that is loosely speaking mapped to infinity by the transformation (division by $x_3$) that reduces the three variables to two.

@@

In contrast to figure \ref{ExampleEllipticCurve} only finite fields $K=GF(p^n)$ are used in elliptic curve cryptography. The reason is loosely speaking that in modern communication engineering data processing is always based on discrete data (simply because computers accept only discrete data).

@@

Let $|E|$ denote the number of elements of an elliptic curve $E$ given an underlying field $GF(k)$ (for practical reasons either $k=p$ with $p$ prim or $k=2^n$). Then Hasse's theorem\cite{ec:Silverman1986} yields $| \, |E| - k-1\,| \le 2\cdot \sqrt{k}$. This Inequality is equivalent to $k+1 - 2\sqrt{k} < |E| < k+1+2\sqrt{k}$. In particular, this means that the number of elements of an elliptic curve is approximately $k$ (for large $k$).

@@

% -----------------------------------------------------------------------------
\section{Security of elliptic-curve-cryptography: The ECDLP}

@@

In elliptic curve cryptography we formally look at points on the elliptic curve as elements of a group with point addition $+$ as operation. Furthermore, we use only elliptic curves that have a sufficiently large number of points. However, in special cases curves may be weak and not useful due to other reasons. For such special cases the ECDLP can be much easier to solve than in the general case. This means that one has to look carefully at the parameters when choosing an elliptic curve for cryptographic applications.

@@

\begin{sloppypar}
  The {\em elliptic curve discrete logarithm
    problem}\index{problem of discrete logarithm} (ECDLP) \index{ECDLP} is the basis for elliptic curve cryptography. Based on this problem, there are different signature schemes. In order to apply one of these, we need:
\end{sloppypar}
\begin{itemize}
    \item An elliptic curve {\bf E} with an underlying field $GF(p^n)$.
    \item A prime $q\ne p$ and a point $G$ on the elliptic curve ${\bf E}$ with order $q$. This means that $qG={\cal O}$ and $rG\ne {\cal O}$ for all $r\in \{1,2,\dots,q-1\}$. Thus $q$ is a factor of the group order (i.e. the number of elements) $\#{\bf E}$ of $E$. Since $q$ is prime, $G$ generates a cyclic sub-group of ${\bf E}$ of order $q$.
\end{itemize}
The parameters mentioned are often called \index{domain parameter}
{\em Domain} parameter. They describe the elliptic curve ${\bf E}$ and 
the cyclic sub-group of ${\bf E}$ on which the signature scheme is based.

@@

\noindent Additionally there is an exhaustive, interactive
\hyperlink{ec:Web-Link:Sage_Massierer}{ECC tutorial} by Maike Massierer.
This interactive introduction to Elliptic Curve Cryptography is built up as
a SageMath notebook.

@@

The ECC notebook\index{elliptic curve!ECC notebook}\index{Massierer, Maike} of
Massierer\footnote{%
Instructions to use an interactive SageMath notebook\index{SageMath!instructions interactive notebook}:\\

@@

\begin{enumerate}
   \setcounter{enumi}{-1}
   \item ECC Notebook (title page and contents)
   \item Introduction and Overview
   \item Motivation for the use of Elliptic Curves in Cryptography
   \item Elliptic Curves in Cryptography
   \item Cryptographic Protocols in ECC
   \item Domain Parameter Generation for ECC Systems
   \item Conclusion and Further Topics
   \item References

% -----------------------------------------------------------------------------
\newpage
\section{Patent aspects}
\index{patent}

@@

A comprehensive description of all implementations and their advantages and disadvantages would go far beyond the scope of this paper. We only want to state that there is a variety of possible implementations for elliptic curve cryptography, much more than for RSA. Therefore, there are serious efforts to reduce this large to a small number of standard implementations. Some standardization committees even try to reduce the complexity by focusing on a small number of (prescribed) curves (ASC-approach).

Today it is still not clear whether these standardization initiatives will be
successful or not. However, without agreed standards, ECC is not likely to
become a real alternative for RSA. The committees might be forced to act fast
if there was a break-through in factorization.

@@

Today elliptic curve cryptography is already in use. A prominent example is the
information network Bonn-Berlin\footnote{%
The Informationsverbund Bonn-Berlin (IVBB) connects governmental institutions
in the old and new German capital.\\
\url{http://www.cio.bund.de/cln_094/sid_92C19118CBA5A021AFD1ABAEC15D2B77/DE/IT-Angebot/IT-Infrastrukturen/IVBB/ivbb_inhalt.html}
                                        }\index{IVBB}, used for the exchange
of strictly confidential documents between different German federal governmental
institutions in Berlin and Bonn. With the help of ECC a high security solution
could be realized. Interoperability, however, played only a minor role.

@@

Both examples show the typical range of application for elliptic curve cryptography: For high security solutions and for implementations on smartcards in which the key length is crucial (because of physical memory available).

%--------------------------------------------------------------------
\newpage
%\addcontentsline{toc}{subsection}{Literaturverzeichnis}
\begin{thebibliography}{99999}
\addcontentsline{toc}{section}{Bibliography}

    \bibitem[Cassels1991]{ec:Cassels1991} J. W. S. Cassels,
	\index{Cassels 1991} \\
        {\em Lectures on elliptic curves},\\
	Cambridge University Press, 1991, 143 pages.
    \bibitem[Koblitz1984]{ec:Koblitz1984}N. Koblitz, 
	\index{Koblitz 1984} \\
        {\em Introduction to elliptic curves and modular forms},\\
	Graduate Texts in Mathematics, Springer-Verlag, 1984.
    \bibitem[Koblitz1998]{ec:Koblitz1998}N. Koblitz,  
	\index{Koblitz 1998}\\
	{\em Algebraic aspects of Cryptography. With an appendix on 
	Hyperelleptic curves by Alfred J. Menezes, Yi Hong Wu and Robert 
	J. Zuccherato}, \\
	Springer-Verlag, 1998, 206 pages.
    \bibitem[Lenstra1987]{ec:Lenstra1987} H.W. Lenstra, 
        \index{Lenstra 1987} \\
        {\em Factoring integers with elliptic curves}, \\
	Annals of Mathematics 126, pp. 649-673, 1987.

\bibitem[Lenstra1999]{ec:Lenstra1999} Arjen K. Lenstra, Eric R. Verheul,
     \index{Lenstra/Verheul 1999} \\
     {\em Selecting Cryptographic Key Sizes (1999 + 2001)}, \\
     Journal of Cryptology 14 (2001), 255-293 \\
     \url{http://www.win.tue.nl/~klenstra/key.pdf},\\
     \url{http://www.cs.ru.nl/E.Verheul/papers/Joc2001/joc2001.pdf},\\
     \url{http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.20.69&rep=rep1&type=pdf}.
     % Former Link is dead: \url{http://www.cryptosavvy.com/cryptosizes.pdf}.

    \bibitem[Menezes1993]{ec:Menezes1993}A. J. Menezes, 
	\index{Menezes 1993} \\
        {\em Elliptic curve public key cryptosystems},\\
	Kluwer Academic Publishers, 1993.
    \bibitem[Silverman1986]{ec:Silverman1986} J. Silverman, 
        \index{Silverman 1986}\\
        {\em The Arithmetic of Elliptic Curves},\\
	Springer-Verlag, 1986.
    \bibitem[Silverman1992]{ec:Silverman1992}J. Silverman,
	\index{Silverman 1992} \\
        {\em The arithmetic of elliptc curves},\\
	Graduate Texts in Mathematics, Springer-Verlag, 1992.
    \bibitem[SilvermanTate1992]{ec:SilvermanTate1992}J. Silverman, J. Tate,
	\index{Silverman/Tate 1992} \\
        {\em Rational points on elliptic curves},\\
	Springer-Verlag, 1992.

\end{thebibliography}

>>>>

\newpage
\hypertarget{ellcurve}{}
\chapter{Elliptic Curves}
\label{Chapter_EllipticCurves}
\index{elliptic curve}
(Filipovics B. / B\"uger M. / Esslinger B. / Oyono R., April 2000,
Updates: Dec. 2001, June 2002, Mar. 2003, Nov. 2009, Aug. 2013, April 2016)

% -----------------------------------------------------------------------------
\section{Elliptic-curve cryptography -- a high-performance substitute for RSA?}
\index{performance}
\label{ECAlternative}

In many business sectors secure and efficient data transfer is essential.
In particular, the RSA algorithm is used in many applications. Although the
security of RSA is beyond doubt, the evolution in computing power has
caused a growth in the necessary key length.  Today, 1024-bit RSA keys are
standard, but the GISA\index{GISA} (German Information Security Agency)
recommends the usage of 2048-bit keys from 2006 on (compare
section~\ref{SecurityRSA}).  The fact that most chips on smart cards cannot
process keys extending 1024 bit shows that there is a need for
alternatives. Elliptic-curve cryptography (ECC) can be such an
alternative in the area of asymmetric cryptography.

@@

In addition, a digital signature can be processed 10-times faster with ECC
than with RSA.  However, verification of a given signature is still more
efficient with RSA than with ECC. Refer to
figure~\ref{ThousandBitMultiplications} (source: Dr.~J.\ Merkle,
Elliptic-Curve Cryptography Workshop, 2001) for a comparison.  The reason is that
RSA public keys can be chosen relatively small as long as the secret key is
long enough.

@@

Although we see no need to replace RSA in any application today\footnote{%
Current information about the security of the RSA algorithm can be found in 
chapter \ref{SecurityRSA}.}, one should
take the usage of ECC-based tools into consideration whenever a new system
is set up --- in particular, when the tool should be available beyond 2005\footnote{%
Compare the recommendation of GISA: ``Fitting Crypto Algorithms'' from October 24th, 2002.
}. More current discussions about the security of ECC can be found in chapter xxxxxxxxxxxxxx

% -----------------------------------------------------------------------------
\section{Elliptic curves -- history}

@@

With a free translation, using the denotation of modern mathematics, this means: \\
No positive whole numbers $x, y$ and $z$ greater than zero exist such tha
t $x^n +y^n = z^n$ for $n>2$. I have found an amazing proof of this fact, but there is
too little space within the confines of this book to include it.

@@

% -----------------------------------------------------------------------------
\section{Elliptic curves -- mathematical basics}

@@

% -----------------------------------------------------------------------------
\subsection{Groups}

@@

% -----------------------------------------------------------------------------
\subsection{Fields}

@@

Fields may contain an infinite number of elements (e.g.\ the field of real numbers).
They are called {\em infinite} fields. In contrast we call a field
{\em finite}, if it contains only a finite number of elements
(e.g.\ ${\mathbb Z}_p = \{0, 1, 2, 3, \cdots, p-1\}$, where $p$
is a prime. ${\mathbb Z}_p$ with addition mod $p$ and multiplication
mod $p$).
\index{field!characteristic}
\paragraph*{Characteristic of a field}
Let $K$ be a field and $1$ be the neutral element of $K$ with
respect to the multiplicative operation ``$\cdot$''. Then the characteristic
of $K$ is said to be the order of $1$ with respect to the additive operation.
This means that the characteristic of $K$ is the smallest positive integer $n$
such that $$ \underbrace{1+1+\dots+1}_{\hbox{$n$ times}} =0 .
$$
If there is no such $n$, i.e. if $1+1+\dots+1\ne 0$ no matter how many $1$s we add, then we call $K$ a field
with characteristic $0$.

@@

In cryptography elliptic curves are a useful tool.\footnote{%
We write ``elliptic-curve cryptography'' with a hyphen, according to ``public-key cryptography''. Sadly, this isn't used consistently in the literature.
}

Such curves are described by some equation. A detailed analysis has shown that curves of the form\footnote{This curve is given by the zeros of a {\it polynomial}\index{polynomial} $F$ of degree three in three variables. In general, expressions of the form
$P=\sum_{i_1,\dots,i_n\in\N_0} a_{i_1\dots i_n} x_1^{i_1}\dots x_n^{i_n}$ with coefficients $a_{i_1\dots i_n}\in K$ are called polynomials in $n$ variables $x_1,\dots,x_n$ with underlying field $K$, if ${\rm deg\,} P:=\max\{i_1+\dots +i_n: a_{i_1\dots i_n}\ne 0\}$ is finite, i.e. the sum has only finitely many non-zero terms (monomials). The sum of the exponents of the variables of each term of the sum is at most $3$, at least one term of the sum has a single variable with $3$ as value of the according exponent.}
\begin{equation}
 F(x_1,x_2,x_3)=-x_1^3+x_2^2x_3+a_1x_1x_2x_3-a_2x_1^2x_3+a_3x_2x_3^2-a_4x_1x_3^2-a_6x_3^3=0,
\label{eccbasisgleichung}
\end{equation}
are especially useful. The variables $x_1,x_2,x_3$ and parameters $a_1,\dots,a_4,a_6$ are elements of a given field $K$, which has certain properties that are make it useful from the cryptographic point of view. The underlying field $K$ might be the well known field of real numbers or some finite field (see last section).
In order to obtain a cure that is useful for cryptography, the parameters have to be chosen in a way that the following conditions hold
$$ \frac{\partial F}{\partial x_1}\ne 0, \quad \frac{\partial F}{\partial x_2}\ne 0, \quad
\frac{\partial F}{\partial x_3}\ne 0 .
$$
We identify points on the curve that can be derived from each over by multiplying each component with some scalar. This makes sense since $(x_1,x_2,x_3)$ solves (\ref{eccbasisgleichung}) if and only if $\alpha (x_1,x_2,x_3)$ ($\alpha\ne 0$) does. Formally, this means that we consider classes of equivalent points instead of single points, where points are called equivalent if one is the scalar multiple of the other one.
\\ If we put $x_3=0$ in the basic equation (\ref{eccbasisgleichung}), then this equation collapses to $-x_1^3=0$, leading to $x_1=0$. Thus, the equivalence class which includes the element $(0,1,0)$ is the only one that contains a point with $x_3=0$. For all points on the curve that are not equivalent to $(0,1,0)$, we may apply the following transformation
$$ K\times K\times (K\setminus\{0\})\ni (x_1,x_2,x_3) \mapsto (x,y):=\left( \frac{x_1}{x_3}, \frac{x_2}{x_3}\right) \in K\times K \, ,
$$
which reduces the number of variables to two instead of three. We note that the
basic equation (\ref{eccbasisgleichung}) $F(x_1,x_2,x_3)=0$  was chosen in a
way that this transformation leads to the famous so-called
Weierstrass-Equation\footnote{Karl Weierstrass\index{Weierstrass, Karl}, 31.10.1815$-$19.12.1897, German
mathematician, famous for his rigorous formal approach to mathematics.} holds
\begin{equation}
 y^2+a_1xy+a_3y = x^3+a_2x^2+a_4x+a_6 \, .
\label{ell}
\end{equation}
Since all but one point (i.e. equivalence class) of the elliptic curve can be described using equation (\ref{ell}), this equation is often called the elliptic equation, and its solutions written as
$$ {\bf E} = \left\{(x,y)\in K\times K \, |\, y^2+a_1xy+a_3y = x^3+a_2x^2+a_4x+a_6  \right\} \cup \{{\cal O} \}.
$$
Here, ${\cal O}$ represents the point $(0,1,0)$ that is loosely speaking mapped to infinity by the transformation (division by $x_3$) that reduces the three variables to two.

@@

In contrast to figure \ref{ExampleEllipticCurve} only finite fields $K=GF(p^n)$ are used in elliptic-curve cryptography. The reason is loosely speaking that in modern communication engineering data processing is always based on discrete data (simply because computers accept only discrete data).

@@

Let $|E|$ denote the number of elements of an elliptic curve $E$ given an underlying field $GF(k)$ (for practical reasons either $k=p$ with $p$ prim or $k=2^n$). Then Hasse's theorem\cite{ec:Silverman1992} yields $| \, |E| - k-1\,| \le 2\cdot \sqrt{k}$. This Inequality is equivalent to $k+1 - 2\sqrt{k} < |E| < k+1+2\sqrt{k}$. In particular, this means that the number of elements of an elliptic curve is approximately $k$ (for large $k$).

@@

% -----------------------------------------------------------------------------
\section{Security of elliptic-curve cryptography: The ECDLP}

@@

In elliptic-curve cryptography we formally look at points on the elliptic curve as elements of a group with point addition $+$ as operation. Furthermore, we use only elliptic curves that have a sufficiently large number of points. However, in special cases curves may be weak and not useful due to other reasons. For such special cases the ECDLP can be much easier to solve than in the general case. This means that one has to look carefully at the parameters when choosing an elliptic curve for cryptographic applications.

@@

\begin{sloppypar}
  The {\em elliptic curve discrete logarithm
    problem}\index{problem of discrete logarithm} (ECDLP) \index{ECDLP} is the basis for elliptic-curve cryptography. Based on this problem, there are different signature schemes. In order to apply one of these, we need:
\end{sloppypar}
\begin{itemize}
    \item An elliptic curve {\bf E} with an underlying field $GF(p^n)$.
    \item A prime $q\ne p$ and a point $G$ on the elliptic curve ${\bf E}$ with order $q$. This means that $qG={\cal O}$ and $rG\ne {\cal O}$ for all $r\in \{1,2,\dots,q-1\}$. Thus $q$ is a factor of the group order (i.e. the number of elements) $\#{\bf E}$ of $E$. Since $q$ is prime, $G$ generates a cyclic sub-group of ${\bf E}$ of order $q$.
\end{itemize}
The parameters mentioned are often called \index{domain parameter}
{\em Domain} parameter. They describe the elliptic curve ${\bf E}$ and 
the cyclic sub-group of ${\bf E}$ on which the signature scheme is based.

@@

\noindent Additionally there is an exhaustive, interactive
\hyperlink{ec:Web-Link:Sage_Massierer}{ECC tutorial} by Maike Massierer.
This interactive introduction to Elliptic-Curve Cryptography is built up as
a SageMath notebook.

@@

The ECC notebook\index{elliptic curve!ECC notebook}\index{Massierer, Maike} of
Massierer\footnote{%
Instructions to use an interactive SageMath notebook\index{SageMath!instructions interactive notebook}:
TODOTODO: Update for the new SageMathCloudxxxxxxxxxxxxx\\

@@

\begin{enumerate}
   \setcounter{enumi}{-1}
   \item ECC Notebook (title page and contents)
   \item Introduction and Overview
   \item Motivation for the use of Elliptic Curves in Cryptography
   \item Elliptic Curves in Cryptography
   \item Cryptographic Protocols in ECC
   \item Domain Parameter Generation for ECC Systems
   \item Conclusion and Further Topics
   \item References
\end{enumerate}

% -----------------------------------------------------------------------------
\newpage
\section{Patent aspects}\index{patent}

@@

A comprehensive description of all implementations and their advantages and disadvantages would go far beyond the scope of this paper. We only want to state that there is a variety of possible implementations for elliptic-curve cryptography, much more than for RSA. Therefore, there are serious efforts to reduce this large to a small number of standard implementations. Some standardization committees even try to reduce the complexity by focusing on a small number of (prescribed) curves (ASC approach).

It is still not clear whether these standardization initiatives will be
successful or not. However, without agreed standards, ECC is not likely to
become a real alternative for RSA. The committees might be forced to act fast
if there was a break-through in factorization.

Current informationen about the patents situation can be found here\footnote{%
\url{https://en.wikipedia.org/wiki/Elliptic_curve_cryptography}\\
\url{https://en.wikipedia.org/wiki/ECC_patents}\\
\url{https://cr.yp.to/ecdh/patents.html} Bernstein, Daniel J. (2006-05-23): ``Irrelevant patents on elliptic-curve cryptography''. Retrieved 2013-08-12.
}.

@@

% Einsatz in Protokollen wie TLS, Chats, ... ergänzen !!  xxxxxxxxxxxxxxxxxxxxxxxxx
Today elliptic-curve cryptography is already broadly in use. A prominent example is the
information network Bonn-Berlin\footnote{%
The Informationsverbund Bonn-Berlin (IVBB) connects governmental institutions
in the old and new German capital.\\
\url{http://www.cio.bund.de/cln_094/sid_92C19118CBA5A021AFD1ABAEC15D2B77/DE/IT-Angebot/IT-Infrastrukturen/IVBB/ivbb_inhalt.html}
                                        }\index{IVBB}, used for the exchange
of strictly confidential documents between different German federal governmental
institutions in Berlin and Bonn. With the help of ECC a high security solution
could be realized. Interoperability, however, played only a minor role.

@@

Both examples show the typical range of application for elliptic-curve cryptography: For high security solutions and for implementations on smartcards in which the key length is crucial (because of lack of physical memory available).

%--------------------------------------------------------------------
\newpage
%\addcontentsline{toc}{subsection}{Literaturverzeichnis}
\begin{thebibliography}{99999}
\addcontentsline{toc}{section}{Bibliography}
    \bibitem[BernsteinLange2013]{ec:BernsteinLange2013} Daniel J. Bernstein, Tanja Lange,
        \index{Bernstein/Lange 2013} \\
        {\em SafeCurves: choosing safe curves for elliptic-curve cryptography}, \\
        \url{http://safecurves.cr.yp.to/}. Retrieved 1 December 2013.
    \bibitem[Cassels1991]{ec:Cassels1991} J. W. S. Cassels,
	\index{Cassels 1991} \\
        {\em Lectures on Elliptic Curves},\\
	Cambridge University Press, 1991, 143 pages.
    \bibitem[Koblitz1984]{ec:Koblitz1984}N. Koblitz, 
	\index{Koblitz 1984} \\
        {\em Introduction to Elliptic Curves and Modular Forms},\\
	Graduate Texts in Mathematics, Springer, 1984.
    \bibitem[Koblitz1998]{ec:Koblitz1998} N. Koblitz,  
	\index{Koblitz 1998}\\
	{\em Algebraic Aspects of Cryptography. With an appendix on 
	Hyperelleptic Curves by Alfred J. Menezes, Yi Hong Wu and Robert 
	J. Zuccherato}, \\
	Springer, 1998, 206 pages.
    \bibitem[Lenstra1987]{ec:Lenstra1987} H.W. Lenstra, 
        \index{Lenstra 1987} \\
        {\em Factoring Integers with Elliptic Curves}, \\
	Annals of Mathematics 126, pp. 649-673, 1987.
    \bibitem[Lenstra1999]{ec:Lenstra1999} Arjen K. Lenstra, Eric R. Verheul,
        \index{Lenstra/Verheul 1999} \\
        {\em Selecting Cryptographic Key Sizes (1999 + 2001)}, \\
        Journal of Cryptology 14 (2001), 255-293 \\
        \url{http://www.win.tue.nl/~klenstra/key.pdf},\\
        \url{http://www.cs.ru.nl/E.Verheul/papers/Joc2001/joc2001.pdf},\\
        \url{http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.20.69&rep=rep1&type=pdf}.
        % Former Link is dead: \url{http://www.cryptosavvy.com/cryptosizes.pdf}.
    \bibitem[Menezes1993]{ec:Menezes1993} A. J. Menezes, 
	\index{Menezes 1993} \\
        {\em Elliptic Curve Public Key Cryptosystems},\\
	Kluwer Academic Publishers, 1993.
    \bibitem[Silverman1992]{ec:Silverman1992} J. Silverman,
	\index{Silverman 1992} \\
        {\em The Arithmetic of Elliptc Curves},\\
	Graduate Texts in Mathematics, Springer, 2nd ed, 2009.
    \bibitem[SilvermanTate1992]{ec:SilvermanTate1992} J. Silverman, J. Tate,
	\index{Silverman/Tate 1992} \\
        {\em Rational Points on Elliptic Curves},\\
	Springer, 1992.
    \bibitem[SchulzWittenEssl2015]{ec:SchulzWittenEssl2015} Ralph-Hardo Schulz, Helmut Witten, and Bernhard Esslinger
	\index{Schulz/Witten/Esslinger 2015} \\
	{\em Rechnen mit Punkten einer elliptischen Kurve}, \\
	LOG IN Journal 181/182 (2015), 13 pages.\\
        \url{http://bscw.schule.de/pub/nj_bscw.cgi/d1024028/Schulz_Witten_Esslinger-Rechnen_mit_Punkten_einer_elliptischen_Kurve.pdf},\\
	Written for teachers, didactically prepared, easy to understand, with many SageMath examples.
	Sadly, this article is only available in German.
\end{thebibliography}
